---
title: "Getting started with quardplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with quardplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 8, dev = "png",
                      dpi = 96, fig.retina = 1)
if (Sys.info()[["sysname"]] == "Linux" && isTRUE(capabilities("cairo"))) options(bitmapType = "cairo")
```

quardplot makes aligned locus figures and tables from GWAS results, a gene
annotation (GFF3 or GTF) and genotypes (VCF or HapMap). This vignette runs the
main steps on a small simulated rice data set. For a longer, step-by-step
guide with your own files, run `quard_guide()`: it copies the guide
`quard_plot_examples.R` into your folder.

## 1. Choose the species

The species gives the chromosome lengths of the reference genome for the
genome-wide plots.

```{r}
library(quardplot)
list_species()[, c("species", "reference_genome", "chromosomes")]
set_species("rice")
```

For a species that is not built in, give its chromosome lengths, for example
`set_species("pepper", chrom_sizes = "pepper_genome.fa.fai")`.

## 2. Input files and checks

`simulate_example_data()` writes files with the same structure as real data:
GWAS results (GAPIT-style CSV), a GFF3 annotation, a VCF and a HapMap file.

```{r}
ex <- simulate_example_data(file.path(tempdir(), "qp_vignette"))
my <- list(gwas = ex$gwas, gff = ex$gff, vcf = ex$vcf, hapmap = ex$complete_hapmap)
check_inputs(my)
```

## 3. The locus figure and tables

```{r, fig.height = 9}
out <- file.path(tempdir(), "qp_vignette_out")
res <- quard_plot(my$gwas, my$gff, my$vcf, chr = ex$chr, pos = ex$pos,
                  output = file.path(out, "locus.pdf"), verbose = FALSE)
res
res$plot
```

The tables are written next to the PDF. The SNP table gives, for every GWAS
SNP in the region, its P value, its r^2^ with the lead SNP and where it lies
relative to the nearest gene:

```{r}
head(res$snp_table[, c("SNP", "POS", "P", "R2_WITH_LEAD", "GENE", "LOCATION", "FEATURE")])
```

Panels and tables are chosen independently, so the LD table can be written
without the heatmap, or tables can be written without any figure:

```{r}
r2 <- quard_plot(my$gwas, my$gff, my$vcf, chr = ex$chr, pos = ex$pos,
                 panels = c("regional", "genes"), tables = c("snps", "ld"),
                 output = file.path(out, "locus_no_heatmap.pdf"), verbose = FALSE)
names(r2$files)
```

With `table_format = "xlsx"` (needs the writexl package) all tables are put in
one Excel file with a README sheet.

## 4. LD and LD blocks

```{r}
g <- genotype_qc(read_vcf(my$vcf, chr = 9, verbose = FALSE), verbose = FALSE)
ld <- calc_ld(g)                       # r2 as in PLINK 1.9 --r2
blocks <- find_ld_blocks(g, verbose = FALSE)   # Gabriel method, as in PLINK --blocks
head(blocks[, c("chr", "start", "end", "kb", "nsnps")])
```

## 5. Where SNPs lie relative to genes

```{r}
a <- annotate_variants(g, my$gff)
table(a$location)
```

## 6. Other tools

```{r, fig.height = 4}
genes <- get_genes(my$gff, chr = 9, pos = ex$pos, flank = 20000, verbose = FALSE)
genes[, c("gene_id", "start", "end", "strand", "distance_bp")]
plot_gene_structure(my$gff, genes$gene_id[1], flank = 2000, variants = my$vcf)
```

```{r, fig.height = 6}
plot_karyotype(my$hapmap, bin_size = 2e6, verbose = FALSE)
```

```{r, fig.height = 7}
tr <- hapmap_tree(my$hapmap, file.path(out, "tree.pdf"), k = 4, verbose = FALSE)
tr$plot
```

## 7. From a terminal

```
Rscript -e "quardplot::quard_cli()" --help
Rscript -e "quardplot::quard_cli()" --species rice --gwas gwas.csv --gff genes.gff3 \
        --vcf genotypes.vcf.gz --target top --out results/locus.pdf
```
